Guidelines for Submissions to ISCA 2013

### 2012-11-21 paper/isca40format/Introduction.tex
before arvind is going to re-org

Large cache-coherent shared-memory multicores are likely to use protocols which
are directory-based, hierarchical, and implemented using packet-switched networks.  Even
though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

@@

The main contributions of this paper are: 1. An automatic procedure to refine a
centralized MSI protocol into protocols for
distributed, hierarchical, cache system with packet-switching; 2. Specification
of the requirements that ensure the correctness of the generated protocol;
3. A discussion of the invariants that simplifies the proof of correctness of
the protocol considerably;
4. Giving a provable lower bound on the number of network virtual channels and
buffers to avoid deadlocks; and 5. Showing that our procedure does not
introduce any overhead over direct implementation of these protocols. We think
that our method of designing protocols substantially reduces the required
effort in implementing provably correct protocols.

\paragraph{Paper organization:} Section \ref{sec:GlobalMsi} describes the MSI
protocol for a hierarchy of caches using nonsuspensive sequential methods. We
use it a specification for proving the correctness of distributed MSI
protocols. Section \ref{sec:DistributedMsi} systematically transforms the
protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
\ref{sec:network} specifies the network, resource and other requirements that
are needed for the distributed MSI protocol. We also give the minimum virtual
channel requirements to ensure the requirements specified in Section
\ref{sec:network}. We show that these formally verified family of protocols
require no extra state bits, nor extra message transfers compared to other
published implementations.  Section \ref{sec:properties} gives a set of
desirable global properties or invariants that are needed to easily establish
the correctness of the distributed protocol.
%It also lists core properties of the distributed protocol that are needed to establish the invariants.
In Section \ref{sec:related}, we present some of the related work and finally
in Section \ref{sec:conclusion}, we present some directions for future work.

>>>>

Large cache-coherent shared-memory multicores are likely to use protocols which
are directory-based, hierarchical, and implemented using packet-switched networks.  Even
though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

@@

The main contributions of this paper are: 1. An automatic procedure to refine a
centralized MSI protocol into protocols for
distributed, hierarchical, cache system with packet-switching; 2. Specification
of the requirements that ensure the correctness of the generated protocol;
3. A discussion of the invariants that simplifies the proof of correctness of
the protocol considerably;
4. Giving a provable lower bound on the number of network virtual channels and
other resources to avoid deadlocks; and 5. Showing that our procedure does not
introduce any overhead over direct implementation of these protocols. We think
that our method of designing protocols substantially reduces the required
effort in implementing provably correct protocols.

\paragraph{Paper organization:} Section \ref{sec:GlobalMsi} describes the MSI
protocol for a hierarchy of caches using nonsuspensive sequential methods. We
use it a specification for proving the correctness of distributed MSI
protocols. Section \ref{sec:DistributedMsi} systematically transforms the
protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
\ref{sec:network} specifies the network, resource and other requirements that
are needed for the distributed MSI protocol. We also give the minimum virtual
channel requirements to ensure the requirements specified in Section
\ref{sec:network}. In Section \ref{sec:implementation}, we show that these formally verified family of protocols
require no extra state bits, nor extra message transfers compared to other
published implementations. Section \ref{sec:properties} gives the local
properties of the distributed protocol that we rely on to easily establish its
correctness, and show how it can be generalized to design other protocols like MOSI.
%It also lists core properties of the distributed protocol that are needed to establish the invariants.
In Section \ref{sec:related}, we present some of the related work and finally
in Section \ref{sec:conclusion}, we present some directions for future work.

### 2012-11-21 paper/isca40format/Invariants.tex
before arvind is going to re-org

\subsection{Local Properties}
\label{sec:localProperties}

@@

The Local-Properties capture the core properties of the distributed MSI
protocol given in Section \ref{sec:DistributedMsi}, and the requirements
mentioned in Section \ref{sec:network}. We will use the violation of
Local-Properties \ref{pSendRespPre} and \ref{cSendRespPre2} as examples to
illustrate the importance of local properties.  In our usual system, let's say
$c_1.state[a] = p.dir[c_1][a] = S$, $c_2.state[a] = p.dir[c_2][a] = I$.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.34]{checkit}
\caption{Effect of violating Invariant \ref{cSendRespPre2} (child evicts a ``pending'' line)}
\label{checkit}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

@@

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.34]{checkit2}
\caption{Effect of violating Invariant \ref{pSendRespPre} (parent sends a voluntary response)}
\label{checkit2}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

@@

In general, any crossing of response messages between a cache and its parent,
like the ones shown in Figures \ref{checkit} and \ref{checkit2} results in
violation of Invariant \ref{conservative}.

>>>>

\section{Local properties that ensure global invariants}
\label{sec:properties}

In this section, we will show some of the core properties that are extracted
from the requirements spelt out in Section \ref{sec:network} as well as the
implementation of the distributed protocol's auxiliary functions given in Figure
\ref{realistic}. These properties can be independently used to prove the
invariants presented in Section \ref{sec:DistributedMsi} (Invariants
\ref{conservative} to \ref{cpSame}). We do not present the proof here, but give
some intuition. Figure \ref{sendReq} lists these local properties.

@@

 We will use the violation of Local-Properties \ref{pSendRespPre} and
\ref{cSendRespPre2} as examples to illustrate the importance of local
properties.  In our usual system, let's say $c_1.state[a] = p.dir[c_1][a] = S$,
$c_2.state[a] = p.dir[c_2][a] = I$.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{checkit}
\caption{Effect of violating Local-Property \ref{cSendRespPre2} (child evicts a ``pending'' line)}
\label{checkit}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

@@

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{checkit2}
\caption{Effect of violating Local-Property \ref{pSendRespPre} (parent sends a voluntary response)}
\label{checkit2}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

@@

In general, any if response messages from both a cache to its parent and from
the parent to the same cache are in flight together, like the scenarios in
Figures \ref{checkit} and \ref{checkit2} Invariant \ref{conservative} will be
violated.

\begin{figure}
\begin{subfigure}{.25\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$O$ & $\rightarrow$ & $S$\\
$S$ & $\rightarrow$ & $O$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\subcaption*{$toCompatible$}
\end{subfigure}
\begin{subfigure}{.25\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $\checkmark$\\
$O$ & $\checkmark$\\
$S$ & $\times$\\
$I$ & $\times$\\
\hline
\end{tabular}
\subcaption*{$dataToUpper$}
\end{subfigure}
\begin{subfigure}{.48\linewidth}
\centering
\begin{tabular}{|c|cccc|}
\hline
& $M$ & $O$ & $S$ & $I$\\
\hline
$M$ & $=$ & $>$ & $>$ & $>$\\
$O$ & $<$ & $=$ & $>$ & $>$\\
$S$ & $<$ & $<$ & $=$ & $>$\\
$I$ & $<$ & $<$ & $<$ & $=$\\
\hline
\end{tabular}
\subcaption*{$<$ (and other) relations}
\end{subfigure}
\caption{MOSI protocol mappings and $<$ relations}
\label{mosi}
\end{figure}

These properties do not require messages to physically come from the sources as
indicated by the requests or responses -- some other cache node can forward
these messages. These can be used to generalize our procedure to
cache-intervention based protocols like MOSI. We generalize the states of MOSI
using the $<$ relation and auxiliary functions $maxCompatible$ and $isModified$,
as shown in Figure \ref{mosi}.

In a hierarchical setting for the MOSI protocol, any cache can send data
with the response granting permissions to the requesting L1 cache, instead of
going through the hierarchy.  On a load request, L1 cache $c$ sends an
upgrade-to-$O$ request, instead of an upgrade-to-$S$ request. The request keeps
going up the hierarchy, sending an upgrade-to-$O$ request to each of the parent
on the way, till it reaches a cache $p$ which has that line in $M$ or $O$
state. If some other child is in $O$ or $M$ state, this node sends a
upgrade-$c$-to-$O$ in addition to a downgrade-to-$S$ request to that cache.
This upgrade-$c$-to-$O$ + downgrade-to-$S$ keeps propogating down the tree till
it reaches a node $n$ whose children are all in a state $< O$. $n$ then
downgrades its state, forwards the upgrade-to-$O$ response to $c$ and
simultaneously sends a fwd-and-downgrade-to-$S$ response to its parent, which
keeps forwarding that till it reaches $p$. If $c$ gets a store request, it will
send an upgrade-to-$M$ request, and $p$ would send a downgrade-to-$I$ request
instead, to its children.

The caches encountered on the way from $n$ to $p$ should handle an extra
scenario where it gets a downgrade response, without a forward response.In this
case, the data and permission forwarding has to be done from this cache to $c$.
It is the designer's responsibility to ensure that this case is handled.

This protocol obeys all the local properties and the requirements mentioned in
this section except that a request sent to a cache might overtake a response
forwarded to that cache. To avoid this, once the forwarding message is
delivered to the intended destination ($c$ in our case), then $c$ sends back an
ACK. $c$'s parent is allowed to to start sending downgrade requests for that
address only after receiving this ACK. This ACK is sent back to each of the
parents till it reaches $p$, which allows each of these caches to send
downgrade requests for that address towards $c$.  This satisfies all the
requirements of Section \ref{sec:network} and the local properties given in
Figure \ref{sendReq}, thereby automatically being correct. The reasoning for
its correctness gets localized to checking if the ordering Requirement
\ref{reqNoOvertakeResp} is maintained.

### 2012-11-21 paper/isca40format/Theorems.tex
before arvind is going to re-org

We first give a list of invariants or global properties that are needed to
ensure that the distributed, hierarchical, MSI protocol in Figure
\ref{realistic} is correct.

\begin{theorem}
It a thread has to send a message, it will eventually be able to send the
message.
\label{canSend}
\end{theorem}

\begin{theorem}
If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
get an upgrade to $y$ response from the parent where $y \ge x$.
\label{willRecv}
\end{theorem}

\begin{theorem}
If a cache sends an downgrade-to-$x$ request to one of its children, it will
eventually get a downgrade-to-$z$ response from the child where $z \le x$.
\label{willRecv2}
\end{theorem}

\begin{theorem}
If a thread is waiting for a line to be allocated, \ie for any address to become
available for eviction, then some address will eventually become available for
eviction
\label{evictDead}
\end{theorem}

\begin{theorem}
If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
just before $p$ sends the response is the same as $c.state[a]$ just before $c$
receives the response.
\label{pcSame}
\end{theorem}

\begin{theorem}
If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
receives the response.
\label{cpSame}
\end{theorem}

\begin{theorem}
Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
p.dir[c][a] \ge c.state[a]$.
\label{conservative}
\end{theorem}

The need for these invariants is intuitive. If, for instance, a thread wants to
send a message in the output channel of the node and is not able to send one
forever, then the thread can not proceed further. If a thread is waiting for a
response (for a request previously sent), and it never gets the response, the
thread can never proceed further. A thread can also not proceed if it's waiting
for being allocated in a free line but never gets one because all the lines are
locked up. These scenarios create a deadlock.

>>>>

We first give a list of invariants or global properties that are needed to
ensure that the distributed, hierarchical, MSI protocol in Figure
\ref{realistic} is correct.
The need for these invariants is intuitive. If, for instance, a thread wants to
send a message in the output channel of the node and is not able to send one
forever, then the thread can not proceed further. If a thread is waiting for a
response (for a request previously sent), and it never gets the response, the
thread can never proceed further. A thread can also not proceed if it's waiting
for being allocated in a free line but never gets one because all the lines are
locked up. These scenarios create a deadlock.

### 2012-11-21 paper/isca40format/stuff/MsiArvind2.tex
before arvind is going to re-org

\section{MSI protocol}
\label{sec:GlobalMsi}

@@

The caches are logically organized in the form of a tree
(Figure~\ref{hierarchy}). The leaves of this tree represent the L1 caches,
with each L1 cache connected to exactly one core. The root of the tree is the
Last-Level-Cache, and its connected to the memory. The cache hierarchy is
inclusive, \ie if an address is present in a cache $c$, then it's also present it its
parent $c.parent$. Note that by address we mean cache-line address, not a byte
address or a word address.

@@

For a non L1 cache, the meaning of coherence states is different. For an
internal cache node $n$, if $n.state[a] = I$, then, the entire sub-tree rooted at $n$
will have coherence state $I$ for address $a$. In particular, no L1 cache in
this subtree will have any permission for $a$. If $n.state[a] = S$, then its
children can either be in $S$ or $I$ state. Finally, if $n.state[a] = M$, then
at most one of its children can be in $M$ state. This guarantees that at most
one of the L1 caches can have write permissions, ensuring the following
invariant:

\begin{theorem}
\textbf{Single-writer Invariant}: When a cache has permission to modify data for
an address, no other cache has read or write permission for that
address.
\label{singleWriter}
\end{theorem}

@@

\begin{theorem}
\textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
the last modified version of the data for the address. If the data for an
address has not been modified, the cache reads the data from memory.
\label{lastRead}
\end{theorem}

@@

We will first describe an atomic version of the hierarchical MSI protocol. Each
of the caches in the hierarchy can be treated like C++/Java objects.
\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}
\small
\begin{algorithmic}
\Proc {\textsc{coreReq}}{$core, a, op, data$}
  \If {$op = Ld$}
    \State \textbf{if} ($this.state[a] < S$)
    \State \;\;\;\; \call{} $this.$\uReq($core, a, S$);
    \State \send{} $this.data[a]$;
  \ElsIf {$op = St$}
    \State \textbf{if} ($this.state[a] < M$)
    \State \;\;\;\; \call{} $this.$\uReq($core, a, M$);
    \State $update(this.data[a], data)$;
  \EndIf
\EndProc
\end{algorithmic}
\caption{L1 cache handling a load or a store request sent by the core by
calling the request handler}
\label{coreHandle}
\end{figure}

In the atomic protocol, a core invokes the L1 cache's request handler
\textsc{coreReq} (Figure \ref{coreHandle}). The L1 cache checks if the request
can already be serviced, if not, it invokes its upgrade request handler \uReq{},
which upgrades the state of the requested address. Finally, in the case of a
load request, the data is sent back to the core, and in the case of a store, the
cache line is updated. Note that the core supplies words or long-words while the
cache line sizes are typically 64 bytes; only the appropriate word in the cache
line is updated in case of a store.

@@

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReq}{$c , a, x$}
  \If {$c.state[a] < x$}
    \If {$a$ is not present in $this$}
      \If {$this$ has no space for $a$}
        \State Choose address $a'$ for eviction;
        \State \textbf{forall} ($c' \in this.children$)
        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
        \State \textbf{if} ($this.state[a'] > I$)
        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
      \EndIf
      \State Allocate a line for $a$;
    \EndIf
    \State // $a$ is already present in $this$
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > toCompatible(x)$)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', toCompatible(x)$);
    \State \textbf{if} ($this$ is Last-Level-Cache \&\& $this.state[a] = I$)
    \State \;\;\;\; \call{} $this$.\sendMem{}($a$);
    \State \textbf{else}
    \State \;\;\;\; \call{} $this.$\uReqL{}($this.parent, a, x$);
  \State \textbf{if} ($c \neq core$)
  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
\label{atomicUReq1}
\end{subfigure}

@@

\caption{MSI protocol}
\label{msi-template}
\end{figure}

Figure \ref{msi-template} gives the methods of a cache to handle the upgrade
requests from its children (or core) and downgrade requests from its parent.
When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
cache $c$ to go to a particular state $x$, $this$ first makes sure that its
other children are compatible with $c$ being in state $x$ for the requested
address. This involves downgrading their states to $toCompatible(x)$. Cache
$this$ has to upgrade its own state before upgrading the state of its child $c$,
if $this$' state is less than the requested state. In this case, $this$ invokes
its parent's handler for upgrade requests. Finally, when state of $this$ is not
less than t$x$ and all its children are compatible with $x$, $this$ can upgrade
$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
in Figure \ref{atomicUReq1}.

A cache, $this$', downgrade handler \dReq{} can also be invoked (for instance
when its parent wants $this$ to downgrade its state to $x$ because the parent's
upgrade request handler was invoked by some other cache. In this case, $this$
makes sure that each of its child's state is not higher than $x$, by invoking
their respective downgrade methods. When all the children have downgraded, the
cache downgrades its own state to $x$, and sends data back to the parent, if it
has been modified (\ie $isModified(x)$ is $True$).

@@

\caption{Methods used in the MSI protocol dealing with other objects}
\label{atomic}
\end{figure}

Figure \ref{msi-template} gives the pseudo-code involved, including the
actions needed to evict a cache line and replace it with another line, in case
there are no free slots available for an address. For reasons that will become
clear in the distributed setting, each cache can directly read or write its own
state and data for each address. It can also directly read it's child's state,
but not directly write it. It can not read it's parents state directly. It can
not directly read or write other cache's data. In order to perform the required
actions indirectly, the cache can invoke other cache's corresponding
\textbf{method}s packaged together as shown in Figure \ref{atomic}. To be
precise we use $this$ explicity even if the object invokes its own methods.

In the case of an L1 cache, once the cache's parent has upgraded its state,
there is no child to upgrade to. Similarly, in the case of a Last-Level-Cache,
the cache simply reads data from the memory if the line is not present in the
cache, since it has no parent whose upgrade request handler it can invoke. All
lines in the Last-Level-Cache are upgraded automatically to be in $M$ state.

\subsection{Refactoring the handler methods}
Figure \ref{msi-template} shows how the request handler methods \uReq{} and
\dReq{} can be refactored to make it amenable to distributed implementation. The
upgrade and downgrade request handlers are not allowed to invoke other cache's
upgrade or downgrade request handlers, or change other cache's states. Instead,
the same calls are made via 5 more local methods listed in Figure \ref{atomic}.
These methods in turn invoke the other caches upgrade and downgrade request
handlers, or change the states of other caches.

@@

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$p, a, x$}
  \State \send{} \Req{this}{p}{a}{x};
  \While {$this.state[a] < x$}
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndWhile
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Upgrading address $a$ of $this$ to $x$, by sending a request to
$this$'s parent $p$}
\label{uReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by sending a
request to $c$}
\label{dReqLocal}
\end{subfigure}

>>>>

\section{MSI protocol}
\label{sec:msi}

@@

The caches are logically organized in the form of a tree
(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
with each L1 cache connected to exactly one core. The root of the tree is the
Last-Level-Cache, and its connected to the memory. The cache hierarchy is
inclusive, \ie if an address is present in a cache $c$, then it's present it its
parent $c.parent$. Note that by address we mean cache-line address, not a byte
address or a word address.

@@

For a non L1 cache, the meaning of coherence states is different. For an
internal node $n$, if $n.state[a] = I$, then, the entire sub-tree rooted at $n$
will have coherence state $I$ for address $a$. In particular, no L1 cache in
this subtree will have any permission for $a$. If $n.state[a] = S$, then its
children can either be in $S$ or $I$ state. Finally, if $n.state[a] = M$, then
at most one of its children can be in $M$ state. This guarantees that at most
one of the L1 caches can have write permissions, ensuring the following
invariant:

\begin{inv}
\textbf{Single-writer Invariant}: When a cache has permission to modify data for
an address, no other cache has read or write permission for that
address.
\label{singleWriter}
\end{inv}

@@

\begin{inv}
\textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
the last modified version of the data for the address. If the data for an
address has not been modified, the cache reads the data from memory.
\label{lastRead}
\end{inv}

@@

We will first describe an atomic version of the hierarchical MSI protocol. In
the atomic protocol, when a core connected to an L1 cache initiates an upgrade
for an address, no other core can initiate an upgrade for any address till the
former L1 cache has finished upgrading. This can be implemented in hardware
using a bus-based system, which locks the bus till an initiated upgrade has
completed.

\floatstyle{boxed}
\restylefloat{figure}

@@

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\textsc{processReq}}{$c, a, op, data$}
  \If {$op = Ld$}
    \State \textbf{if} ($c.state[a] \ge S$)
    \State \;\;\;\; \call{} $c.$\uReq($this, a, S$);
    \State \textbf{return} $c.data[a]$;
  \ElsIf {$op = St$}
    \State \textbf{if} ($c.state[a] = M$)
    \State \;\;\;\; \call{} $c.$\uReq($this, a, M$);
    \State $update(c.data[a], data)$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{core handling a load or a store request by calling the request
handler in its L1 cache}
\label{coreHandle}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReq}{$c , a, x$}
    \If {$a$ is not present in $this$}
      \If {$this$ has no space for $a$}
        \State Choose address $a'$ for eviction;
        \State \textbf{forall} ($c' \in this.children$)
        \State \;\;\;\; \textbf{if} ($c'.state[a'] > I$)
        \State \;\;\;\;\;\;\;\; \call{} $c'.$\dReq($this, a', I$);
        \If {$this.state[a'] > I$}
          \State \textbf{if} ($this.state[a'] = M$)
          \State \;\;\;\; $(this.parent).data[a'] \gets this.data[a']$;
          \State $this.state[a'] \gets I$;
        \EndIf
      \EndIf
      \State Allocate a line for $a$;
    \EndIf
    \State // $a$ is already present in $this$
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > toCompatible(x)$)
    \State \;\;\;\;\;\;\;\; \call{} $c'.$\dReq($this, a', toCompatible(x)$);
    \State \textbf{if} ($this$ is Last-Level-Cache \&\& $this.state[a] = I$) \bopen
    \State \;\;\;\; $this.data[a] \gets memory[a]$;
    \State \;\;\;\; $this.state[a] \gets M$;
    \State \bclose \;\textbf{else}
    \State \;\;\;\; \call{} $(this.parent.)$\uReq{}($this, a, x$);
  \If {$c \neq core$}
  \State \textbf{if} ($c.state[a] = I$)
  \State \;\;\;\; $c.data[a] \gets this.data[a]$;
  \State $c.state[a] \gets x$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
\label{atomicUReq1}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReq}{$p , a, x$}
  \If {$this.state[a] > x$}
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
    \State \;\;\;\;\;\;\;\; \call{} $c'.$\dReq($this, a, x$);
    \State \textbf{if} ($this.state[a] = M$)
    \State \;\;\;\; $p.data[a] \gets this.data[a]$;
    \State $this.state[a] \gets x$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's downgrade request handler -- request from $this$' parent, $p$, to upgrade $this$' state to $x$ for address $a$ }
\label{atomicDReq1}
\end{subfigure}

\caption{Atomic MSI protocol}
\label{msi-simple}
\end{figure}

Figure \ref{msi-simple} gives the pseudo-code for handling a request from a
core. The caches and the cores can be thought of as objects with methods. The
core has a \textsc{processReq} method which it invokes to process a load or
store request. The core first checks if its L1 cache has the required state. If
not, it invokes the L1 cache's upgrade request handler \uReq{}. \uReq{} in turn
calls the upgrade request handler of which would upgrade the state of the
L1-cache, and finally the request from the core will be serviced. Note that the
core supplies words or long-words while the cache line sizes are typically 64
bytes; only the appropriate word in the cache line is updated in case of a
store.

In the case of a non-L1 cache, $this$, when its upgrade request handler is
invoked by its child $c$, if the requested addres is not already in $this$, it
tries to get a line for the address. If the cache has no empty slots for the
requested address, an existing address is evicted. In order to evict an address
$a'$, to maintain the inclusiveness property, the address $a'$ has to be evicted
in all of the $this$' children. This is done by calling the downgrade request
handler \dReq{} of each of its child which is at a higher state than $I$ for
that address; such a child is asked to downgrade to $I$ state. Once the children
have evicted, if data in $this$ is modified with respect to its parent, then the
data is written back. At this point a line is allocated for the originally
requested address.

Once $this$ has established that it has a line for the requested address, it now
has to ensure that the states of its other children are compatible with the
requested upgrade; $this$ again invokes its children's downgrade request handler
\dReq{}, for all the children who are at a higher state than what is compatible,
to downgrade them to the compatible state. $this$ invokes its parents upgrade
request handler ($(this.parent).\uReq{}$), which in turn will will do the same
sequence of operations and eventually upgrade $this$' state. Finally, $this$
upgrades the state of the requesting child $c$, and sends data to $c$ if
necessary.

When a cache, $this$', downgrade handler is invoked by its parent, it in turn
invokes the downgrade handler of all its children which are at a higher state
than what the parent requested. Once all the children have downgraded their
respective states, $this$ downgrades its own state and writes the data back to
its parent if it has been modified with respect to the parent, \ie if $this$ is
in $M$ state.

In the case of an L1 cache, once the cache's parent has upgraded its state,
there is no child to upgrade to. Similarly, in the case of a Last-Level-Cache,
the cache simply reads data from the memory if the line is not present in the
cache, since it has no parent whose upgrade request handler it can invoke. All
lines in the Last-Level-Cache are made to be in $M$ state.

\subsection{Refactoring the handler methods}
\begin{figure}
\small

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReq}{$c , a, x$}
    \If {$a$ is not present in $this$}
      \If {$this$ has no space for $a$}
        \State Choose address $a'$ for eviction;
        \State \textbf{forall} ($c' \in this.children$)
        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
        \State \textbf{if} ($this.state[a'] > I$)
        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
      \EndIf
      \State Allocate a line for $a$;
    \EndIf
    \State // $a$ is already present in $this$
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > toCompatible(x)$)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', toCompatible(x)$);
    \State \textbf{if} ($this$ is Last-Level-Cache \&\& $this.state[a] = I$)
    \State \;\;\;\; \call{} $this$.\sendMem{}($a$);
    \State \textbf{else}
    \State \;\;\;\; \call{} $this.$\uReqL{}($this.parent, a, x$);
  \State \textbf{if} ($c \neq core$)
  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
\label{atomicUReq1}
\end{subfigure}

@@

\caption{MSI protocol -- general template}
\label{msi-template}
\end{figure}

@@

\caption{Atomic implementation of MSI protocol}
\label{atomic}
\end{figure}

Figure \ref{msi-template} shows how the request handler methods \uReq{} and
\dReq{} can be refactored to make it amenable to distributed implementation. The
upgrade and downgrade request handlers are not allowed to invoke other cache's
upgrade or downgrade request handlers, or change other cache's states. Instead,
the same calls are made via 5 more local methods listed in Figure \ref{atomic}.
These methods in turn invoke the other caches upgrade and downgrade request
handlers, or change the states of other caches.

@@

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$p, a, x$}
  \State \send{} \Req{this}{p}{a}{x};
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Upgrading address $a$ of $this$ to $x$, by sending a request to
$this$'s parent $p$}
\label{uReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by sending a
request to $c$}
\label{dReqLocal}
\end{subfigure}
